Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-07-27 versions/response.v1.txt

Thank you for bringing this to our attention.
We have tried to use branch lengths in as many instances as we could in the discussions, using fidelities only in the computations as they allow for great simplifications.

>  - Fig 5: it is surprising that \hat{w} seems completely flat as a
>  function of x* until it jumps discontinuously to 0. Why not include
>  the analytical bound in Fig 5?

The discontinuity is a reflection of the change in ancestral states when the generating parameters change.
As the true generating parameters change, we become more likely to choose a different ancestral state as most likely, resulting in an entirely different objective function to be optimized.
We have added the following:
"There are discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}) due to the choice of which ancestral state splits are maximal."

The analytical bound is a coarse estimate of where inconsistency arises.
In the revised manuscript, we fix the maximal ancestral state to simplify the argument, and comparing the analytical with the empirical would not provide much added benefit to the analysis.

@@

>  - It would be nice if the abstract pointed out that these
>  inconsistencies arise when branch lengths are long and that the
>  approximate joint maximization procedure was suggested for precisely
>  the opposite limit. Similar at the beginning of the discussion, I
>  think it would be more appropriate if this would say smth along the
>  lines "Sagulenko et al suggested an approximation for short
>  branches... we show that this doesn't work for long branches..."

TODO we should do some change, but make it clear that it's consistently biased even for short branches.
%das: I believe this is done
We have added the line:
"Although the software described in \citet{Sagulenko2018-xl} fits on a user-supplied topology and the authors explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference."

@@

> However, although my original intent was to carefully work through the
> math (including the derivations and proofs in the appendix), I have
> not found the time to do so, and because of the lateness of my review
> at this point, I will provide just a few comments in the hope that
> they will be useful.  Hopefully, other reviewers were able to check
> the correctness of the mathematical results more carefully.

Thank you for your assessment.
We have worked hard in our initial submission and in our revision to write out the argument in great (indeed, excruciating) detail, and we would appreciate it if you could take a look at the revision.

> The only real criticism I have of the paper is that the authors could
> have done a better job in connecting their new work to previous
> results.  For example, Goldman (1990, Syst. Biol. 39:345-361) showed
> that joint estimation of topology and ancestral states by maximum
> likelihood was equivalent to maximum parsimony analysis for the same
> 2-state model considered by Shaw and Matsen when all branches are
> forced to have the same length.  They noted the statistical
> inconsistency of this approach, which was necessarily true because
> parsimony was already known to be inconsistent (the famous result of
> Felsenstein 1978).  Shaw and Matsen extend this result to the case of
> unequal branch lengths, which while not particularly surprising, is
> nonetheless an important extension.  The authors should also cite key
> relevant earlier work on the identifiability ot the tree topology with
> "classical" maximum likelihood phylogenetic estimation (Allman et al
> 2008, Adv Appl Prob 40:229–249; Chai and Housworth 2011, Syst Biol
> 60:713–718) in addition to crediting Roychoudhury 2015 for
> establishing ML consistency.

We are very grateful to this reviewer for pointing out the Goldman paper.
We were not aware of this paper and now have a paragraph describing it and comparing to our work.
See the fourth paragraph of the Discussion section.
Additionally, we have used his terminology of "integrated likelihood" to avoid confusion.

It's also a good idea to cite additional papers about identifiability, and we have done so in the Introduction:
"An estimator is said to be statistically consistent if it converges to the generating model with probability one in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \citep{Allman2008-wd,Chai2011-ff,RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood."
and Discussion sections:
"Consistency proofs of standard maximum likelihood estimates of phylogeny \eqref{eq:marginal_likelihood} are recent \citep{Allman2008-wd,Chai2011-ff,RoyChoudhury2015-ta}, and no results have been obtained for profile likelihood."
